---
title: "Context mapping"
subtitle: "Inputs behind the action map — not a second prioritization"
format:
  html:
    toc: false
    page-layout: full
    css: styles.css
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
library(sf)
library(leaflet)
library(dplyr)

hex_path <- "outputs/hex/faa_hex_scores.geojson"
has_hex <- file.exists(hex_path)

bin_levels <- c("Very Low", "Low", "Moderate", "High", "Very High")

# White → dark red (Wildfire Exposure Risk category)
wfe_cat_colors <- c(
  "Very Low" = "#ffffff",
  "Low" = "#ffffff",
  "Moderate" = "#fdd49e",
  "High" = "#e34a33",
  "Very High" = "#a50f15"
)

# Teal (AOI-relative building density = PEOPLE_CAT; VL/Low white so sparse AOI reads light)
people_cat_colors <- c(
  "Very Low" = "#ffffff",
  "Low" = "#ffffff",
  "Moderate" = "#99d8c9",
  "High" = "#41ae76",
  "Very High" = "#005824"
)

# Soft purple (AOI-relative Census housing density — context only)
housing_cat_colors <- c(
  "Very Low" = "#ffffff",
  "Low" = "#ffffff",
  "Moderate" = "#dadaeb",
  "High" = "#bcbddc",
  "Very High" = "#756bb1"
)

# Layer control + legend must use the same string (sync JS matches on it)
people_layer <- "Relative building density (AOI)"
housing_layer <- "Relative housing density (Census units)"

pine_codes <- tryCatch(
  {
    read.csv("config/evt_pine_barrens.csv", stringsAsFactors = FALSE)$evt_code |>
      as.character() |>
      trimws()
  },
  error = function(e) character()
)

# AOI quintile bins (same cut logic as PEOPLE_CAT in script 04)
aoi_quintile_bin <- function(x) {
  x <- as.numeric(x)
  x[is.na(x)] <- 0
  qs <- as.numeric(stats::quantile(x, probs = c(0.2, 0.4, 0.6, 0.8), na.rm = TRUE, type = 7))
  out <- rep("Very High", length(x))
  out[x < qs[1]] <- "Very Low"
  out[x >= qs[1] & x < qs[2]] <- "Low"
  out[x >= qs[2] & x < qs[3]] <- "Moderate"
  out[x >= qs[3] & x < qs[4]] <- "High"
  factor(out, levels = bin_levels)
}
```

These layers are **context**: they help interpret why a hex got an action. They do **not** replace the [Actions](index.qmd) map. Switch layers with the control (top-right); the legend updates with the active layer.

**Building density** (teal) is the model's **proxy for people** — WRTC footprints ≥40 m² (homes, cabins, camps, commercial, industrial), so it picks up structures that Census housing undersells. It shows where structures are, not how many people live, work or recreate there. **Housing density** (purple) is map-only: Census housing units, including seasonal / vacation homes. It does **not** pick actions or Goldilocks. Both are **relative to this Northwoods AOI** — High / Very High means “more *here* than elsewhere in the AOI,” not urban density.

PAD is shown for orientation only — it does not pick actions or Goldilocks scores. See [Methods](methods.qmd).

```{r}
#| label: context-map
#| fig-height: 9
if (!has_hex) {
  cat(
    "**No hex output yet.** Run the Pro pipeline through `05_export_hex_geojson.py`, ",
    "commit `outputs/hex/faa_hex_scores.geojson`, then re-render.\n"
  )
} else {
  hex <- st_read(hex_path, quiet = TRUE)
  has_building <- "WRTC_BLDG_DENSITY_MEAN" %in% names(hex) &&
    any(!is.na(as.numeric(hex$WRTC_BLDG_DENSITY_MEAN)))
  has_housing <- "WRTC_HU_DENSITY_MEAN" %in% names(hex) &&
    any(!is.na(as.numeric(hex$WRTC_HU_DENSITY_MEAN)))

  if (!has_building) {
    cat(
      "> **Building density not on hexes yet.** Set `wrtc_building_density` in ",
      "`config/paths.local.yaml`, re-run `02` → `05`, then re-render. Until then ",
      "the people layer shows bins from an older housing-based export.\n\n"
    )
  }

  # Pine/oak amount = sum of top-3 EVT area % on the pine/oak list
  evt_code <- function(v) {
    ifelse(is.na(v) | v == "", NA_character_, as.character(as.integer(as.numeric(v))))
  }
  pct_if_pine <- function(code, pct) {
    code <- evt_code(code)
    pct <- as.numeric(pct)
    ifelse(!is.na(code) & code %in% pine_codes & !is.na(pct), pct, 0)
  }

  hex <- hex |>
    mutate(
      WFE_CAT = factor(as.character(WFE_CAT), levels = bin_levels),
      PEOPLE_CAT = factor(
        as.character(PEOPLE_CAT),
        levels = bin_levels
      ),
      PADUS_FRAC = as.numeric(PADUS_FRAC),
      FDIST_FUEL_DELTA = as.numeric(FDIST_FUEL_DELTA),
      PINE_OAK_PCT = pct_if_pine(EVT_1, EVT_1_PCT) +
        pct_if_pine(EVT_2, EVT_2_PCT) +
        pct_if_pine(EVT_3, EVT_3_PCT),
      PINE_OAK_PCT = pmin(100, PINE_OAK_PCT)
    )

  if (has_housing) {
    hex <- hex |>
      mutate(
        WRTC_HU_DENSITY_MEAN = as.numeric(WRTC_HU_DENSITY_MEAN),
        HOUSING_CAT = aoi_quintile_bin(WRTC_HU_DENSITY_MEAN)
      )
  }

  wfe_cat_pal <- colorFactor(
    palette = unname(wfe_cat_colors[bin_levels]),
    domain = bin_levels,
    levels = bin_levels,
    ordered = TRUE,
    na.color = "#cccccc"
  )
  people_cat_pal <- colorFactor(
    palette = unname(people_cat_colors[bin_levels]),
    domain = bin_levels,
    levels = bin_levels,
    ordered = TRUE,
    na.color = "#cccccc"
  )
  pad_pal <- colorNumeric(
    palette = c("#ffffff", "#a1d99b", "#41ab5d", "#006d2c", "#00441b"),
    domain = c(0, 1),
    na.color = "#cccccc"
  )
  pine_pal <- colorNumeric(
    palette = c("#ffffff", "#a1d99b", "#238b45", "#00441b"),
    domain = c(0, 100),
    na.color = "#cccccc"
  )
  # Diverging: fuel remove (−, green) → neutral → fuel add (+, brown)
  fdist_lim <- max(0.25, max(abs(hex$FDIST_FUEL_DELTA), na.rm = TRUE))
  fdist_pal <- colorNumeric(
    palette = c("#1b7837", "#f5f5f5", "#8c510a"),
    domain = c(-fdist_lim, fdist_lim),
    na.color = "#cccccc"
  )

  evt_bit <- if ("EVT_NAME" %in% names(hex)) {
    ifelse(
      !is.na(hex$EVT_NAME) & hex$EVT_NAME != "",
      paste0("Majority EVT: ", hex$EVT_NAME, "<br/>"),
      ""
    )
  } else {
    rep("", nrow(hex))
  }
  peat_bit <- if ("PEAT_HEX" %in% names(hex)) {
    paste0("Wetland / peat flag: ", hex$PEAT_HEX)
  } else {
    rep("", nrow(hex))
  }
  housing_bit <- if (has_housing) {
    paste0(
      "Relative housing density (context): ", as.character(hex$HOUSING_CAT), "<br/>"
    )
  } else {
    rep("", nrow(hex))
  }

  hex <- hex |>
    mutate(
      fdist_dominate = dplyr::case_when(
        is.na(FDIST_FUEL_DELTA) ~ "n/a",
        FDIST_FUEL_DELTA >= 0.05 ~ "dominated by fuel add",
        FDIST_FUEL_DELTA <= -0.05 ~ "dominated by fuel remove",
        TRUE ~ "mixed / near neutral"
      ),
      popup_html = paste0(
        "<b>", GRID_ID, "</b><br/>",
        "Wildfire Exposure Risk: ", as.character(WFE_CAT), "<br/>",
        "Relative building density (people proxy): ", as.character(PEOPLE_CAT), "<br/>",
        housing_bit,
        "Pine/oak in top 3 EVTs: ", round(PINE_OAK_PCT, 0), "% of hex<br/>",
        "PAD-US GAP 1–3 fraction: ", round(PADUS_FRAC, 2), "<br/>",
        "Fuel change (FDist): ", round(FDIST_FUEL_DELTA, 3),
        " — ", fdist_dominate, "<br/>",
        evt_bit,
        peat_bit
      )
    )

  bb <- st_bbox(hex)
  pad_x <- (bb[["xmax"]] - bb[["xmin"]]) * 0.01
  pad_y <- (bb[["ymax"]] - bb[["ymin"]]) * 0.01
  west <- bb[["xmin"]] - pad_x
  south <- bb[["ymin"]] - pad_y
  east <- bb[["xmax"]] + pad_x
  north <- bb[["ymax"]] + pad_y

  base_groups <- c(
    "Wildfire Exposure Risk category",
    people_layer,
    if (has_housing) housing_layer else NULL,
    "Pine/oak amount (top 3 EVTs)",
    "PAD-US GAP 1–3 fraction",
    "Recent fuel change (FDist)"
  )

  m <- leaflet(
    height = "85vh",
    width = "100%",
    options = leafletOptions(zoomSnap = 0.25, zoomDelta = 0.5)
  ) |>
    # Esri gray canvas: light underlays without CARTO API-key watermarks
    addProviderTiles("Esri.WorldGrayCanvas") |>
    fitBounds(west, south, east, north) |>
    addPolygons(
      data = hex,
      fillColor = ~ unname(wfe_cat_colors[as.character(WFE_CAT)]),
      fillOpacity = 0.75,
      color = "#333333",
      weight = 0.1,
      opacity = 0.35,
      group = "Wildfire Exposure Risk category",
      popup = ~ popup_html
    ) |>
    addPolygons(
      data = hex,
      fillColor = ~ unname(people_cat_colors[as.character(PEOPLE_CAT)]),
      fillOpacity = 0.75,
      color = "#333333",
      weight = 0.1,
      opacity = 0.35,
      group = people_layer,
      popup = ~ popup_html
    )

  if (has_housing) {
    m <- m |>
      addPolygons(
        data = hex,
        fillColor = ~ unname(housing_cat_colors[as.character(HOUSING_CAT)]),
        fillOpacity = 0.75,
        color = "#333333",
        weight = 0.1,
        opacity = 0.35,
        group = housing_layer,
        popup = ~ popup_html
      )
  }

  m <- m |>
    addPolygons(
      data = hex,
      fillColor = ~ pine_pal(PINE_OAK_PCT),
      fillOpacity = 0.75,
      color = "#333333",
      weight = 0.1,
      opacity = 0.35,
      group = "Pine/oak amount (top 3 EVTs)",
      popup = ~ popup_html
    ) |>
    addPolygons(
      data = hex,
      fillColor = ~ pad_pal(PADUS_FRAC),
      fillOpacity = 0.75,
      color = "#333333",
      weight = 0.1,
      opacity = 0.35,
      group = "PAD-US GAP 1–3 fraction",
      popup = ~ popup_html
    ) |>
    addPolygons(
      data = hex,
      fillColor = ~ fdist_pal(FDIST_FUEL_DELTA),
      fillOpacity = 0.75,
      color = "#333333",
      weight = 0.1,
      opacity = 0.35,
      group = "Recent fuel change (FDist)",
      popup = ~ popup_html
    ) |>
    addLegend(
      position = "bottomright",
      colors = unname(wfe_cat_colors[bin_levels]),
      labels = bin_levels,
      title = "Wildfire Exposure Risk",
      opacity = 0.9,
      group = "Wildfire Exposure Risk category"
    ) |>
    addLegend(
      position = "bottomright",
      colors = unname(people_cat_colors[bin_levels]),
      labels = bin_levels,
      title = people_layer,
      opacity = 0.9,
      group = people_layer
    )

  if (has_housing) {
    m <- m |>
      addLegend(
        position = "bottomright",
        colors = unname(housing_cat_colors[bin_levels]),
        labels = bin_levels,
        title = housing_layer,
        opacity = 0.9,
        group = housing_layer
      )
  }

  title_map_extra <- if (has_housing) {
    sprintf(",\n          '%s': '%s'", housing_layer, housing_layer)
  } else {
    ""
  }

  m <- m |>
    addLegend(
      position = "bottomright",
      pal = pine_pal,
      values = c(0, 100),
      title = "Pine/oak % (top 3 EVTs)",
      opacity = 0.9,
      group = "Pine/oak amount (top 3 EVTs)"
    ) |>
    addLegend(
      position = "bottomright",
      pal = pad_pal,
      values = c(0, 1),
      title = "PAD-US GAP 1–3 fraction",
      opacity = 0.9,
      group = "PAD-US GAP 1–3 fraction"
    ) |>
    addLegend(
      position = "bottomright",
      pal = fdist_pal,
      values = c(-fdist_lim, fdist_lim),
      title = "Fuel (− remove / + add)",
      opacity = 0.9,
      group = "Recent fuel change (FDist)"
    ) |>
    addLayersControl(
      baseGroups = base_groups,
      options = layersControlOptions(collapsed = FALSE)
    ) |>
    htmlwidgets::onRender(sprintf(
      "
      function(el, x) {
        var map = this;
        var bounds = [[%f, %f], [%f, %f]];
        var titleMap = {
          'Wildfire Exposure Risk category': 'Wildfire Exposure Risk',
          '%s': '%s'%s,
          'Pine/oak amount (top 3 EVTs)': 'Pine/oak',
          'PAD-US GAP 1–3 fraction': 'PAD-US GAP 1–3 fraction',
          'Recent fuel change (FDist)': 'Fuel'
        };
        function legendTitle(leg) {
          var s = leg.querySelector('strong');
          return s ? s.textContent.trim() : '';
        }
        function syncLegends() {
          var legends = el.querySelectorAll('.legend');
          var inputs = el.querySelectorAll('.leaflet-control-layers-base input');
          var active = null;
          inputs.forEach(function(inp) {
            if (inp.checked) active = inp.parentElement.textContent.trim();
          });
          var want = titleMap[active] || '';
          legends.forEach(function(leg) {
            var title = legendTitle(leg);
            leg.style.display = (want && title.indexOf(want) === 0) ? 'block' : 'none';
          });
        }
        function fitAOI() {
          map.invalidateSize(false);
          map.fitBounds(bounds, { padding: [10, 10], animate: false });
        }
        map.on('baselayerchange', syncLegends);
        setTimeout(syncLegends, 50);
        var nFit = 0;
        function fitOnce() { fitAOI(); nFit += 1; }
        fitOnce();
        setTimeout(fitOnce, 150);
        setTimeout(fitOnce, 500);
        setTimeout(fitOnce, 1200);
        if (window.ResizeObserver) {
          var ro = new ResizeObserver(function() {
            if (nFit < 8) fitOnce(); else ro.disconnect();
          });
          ro.observe(el);
        }
      }
      ",
      south, west, north, east,
      people_layer, people_layer,
      title_map_extra
    ))

  m
}
```
