---
author: "Randy Swaty"
title: "Actions and Priority Areas: Fire Adapted Ecosystems Focus"
subtitle: "Setting the stage for conversation and deeper dives"
format:
  html:
    toc: false
    page-layout: full
    css: styles.css
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
library(sf)
library(leaflet)
library(dplyr)

hex_path <- "../outputs/hex/faa_hex_scores.geojson"
has_hex <- file.exists(hex_path)

action_levels <- c(
  "Value to protect from fire",
  "Treat fire risk for people",
  "Ecosystem health focus",
  "Wetlands — assess locally",
  "Defer and Monitor"
)
action_colors <- c(
  "Value to protect from fire" = "#ba620f",
  "Treat fire risk for people" = "#5c184b",
  "Ecosystem health focus" = "#0f451d",
  "Wetlands — assess locally" = "#b0b6f0",  # lighter than other actions
  "Defer and Monitor" = "#74cc8c"
)

# Nested people-first bands among non-defer hexes (5% darkest).
# Outside-band hexes are not drawn (see map) — pale fills were hard to see in WI.
gold_levels <- c("Top 5%", "Top 10%", "Top 15%")
gold_colors <- c(
  "Top 5%" = "#00441b",
  "Top 10%" = "#238b45",
  "Top 15%" = "#66c2a4"
)
```

The Northwoods Region has no shortage of fire adapted ecosystems.  We have calculated that at least 4.6M acres are in moderate to high Wildfire Exposure Risk categories (see [initial exploration (Risk tab)](https://rswaty.github.io/northwoods/)). In this dashboard I work to prioritize what to do where with aggressive filtering, housing density data and draft "Action" assignments in hopes of getting at actionable areas. I am trying to find the middle ground, aka "The Goldilocks" range of areas-a stretch while also being doable.  **These maps are not meant to be the final answer** but instead guides of where to dig deeper.


You can toggle **Action class** vs **Goldilocks** in the top-right panel. For more information also see [Context](context.qmd) inputs and exploratory datasets and [Methods](methods.qmd) for details of how these maps were made.

```{r}
#| label: map
#| fig-height: 9
if (!has_hex) {
  cat(
    "**No hex output yet.** Run the Pro pipeline through `05_export_hex_geojson.py`, ",
    "commit `outputs/hex/faa_hex_scores.geojson`, then re-render this dashboard.\n"
  )
} else {
  hex <- st_read(hex_path, quiet = TRUE) |>
    mutate(
      action_label = case_match(
        ACTION_CLASS,
        "value_to_protect_from_fire" ~ "Value to protect from fire",
        "protect_from_fire" ~ "Value to protect from fire",  # legacy id
        "treat_fire_risk_for_people" ~ "Treat fire risk for people",
        "ecosystem_health_focus" ~ "Ecosystem health focus",
        "wetlands_assess_locally" ~ "Wetlands — assess locally",
        "defer_monitor" ~ "Defer and Monitor",
        .default = "Defer and Monitor"
      ),
      action_label = factor(action_label, levels = action_levels),
      gold_pri = as.integer(round(as.numeric(GOLDILOCKS_PRIORITY))),
      # Wetlands stay on the action map but are not in Goldilocks bands
      gold_label = case_when(
        ACTION_CLASS == "wetlands_assess_locally" ~ NA_character_,
        gold_pri == 3L ~ "Top 5%",
        gold_pri == 2L ~ "Top 10%",
        gold_pri == 1L ~ "Top 15%",
        TRUE ~ NA_character_
      ),
      gold_label = factor(gold_label, levels = gold_levels),
      is_wetland = ACTION_CLASS == "wetlands_assess_locally"
    )

  # Explicit levels= so leaflet/scales does not alphabetize and scramble colors.
  action_pal <- colorFactor(
    palette = unname(action_colors[action_levels]),
    domain = action_levels,
    levels = action_levels,
    ordered = TRUE,
    na.color = "#9D9D9D"
  )
  gold_pal <- colorFactor(
    palette = unname(gold_colors[gold_levels]),
    domain = gold_levels,
    levels = gold_levels,
    ordered = TRUE,
    na.color = "#eeeeee"
  )

  # People-first urgency on a 0–100 scale for the popup (100 = highest in this AOI)
  sp <- hex$SCORE_PEOPLE
  sp_min <- min(sp, na.rm = TRUE)
  sp_max <- max(sp, na.rm = TRUE)
  hex <- hex |>
    mutate(
      urgency_0_100 = if (is.finite(sp_max) && sp_max > sp_min) {
        round(100 * (SCORE_PEOPLE - sp_min) / (sp_max - sp_min), 0)
      } else {
        0
      }
    )

  # Precompute EVT top-3 popup block (area ranks from script 03)
  if (all(c("EVT_1_NAME", "EVT_1_PCT") %in% names(hex))) {
    fmt_one <- function(nm, pct, rank) {
      ifelse(
        is.na(nm) | nm == "",
        "",
        paste0(
          rank, ". ", nm, " (",
          ifelse(is.na(pct), "?", paste0(round(pct, 0), "%")),
          ")"
        )
      )
    }
    e1 <- fmt_one(hex$EVT_1_NAME, hex$EVT_1_PCT, 1)
    e2 <- if ("EVT_2_NAME" %in% names(hex)) {
      fmt_one(hex$EVT_2_NAME, hex$EVT_2_PCT, 2)
    } else {
      rep("", nrow(hex))
    }
    e3 <- if ("EVT_3_NAME" %in% names(hex)) {
      fmt_one(hex$EVT_3_NAME, hex$EVT_3_PCT, 3)
    } else {
      rep("", nrow(hex))
    }
    hex$EVT_TOP3_HTML <- vapply(seq_len(nrow(hex)), function(i) {
      parts <- c(e1[i], e2[i], e3[i])
      parts <- parts[nzchar(parts)]
      if (!length(parts)) {
        return("")
      }
      paste0(
        "<br/><b>EVT (top 3 by area)</b><br/>",
        paste(parts, collapse = "<br/>"),
        "<br/><i>WFE is blended ember transmission across the hex fuel mosaic (Beverly-style), not majority EVT alone.</i>"
      )
    }, character(1))
  } else if ("EVT_NAME" %in% names(hex)) {
    hex$EVT_TOP3_HTML <- ifelse(
      is.na(hex$EVT_NAME) | hex$EVT_NAME == "",
      "",
      paste0("<br/>EVT: ", hex$EVT_NAME)
    )
  } else {
    hex$EVT_TOP3_HTML <- ""
  }

  # Goldilocks layer: in-band hexes only (after popup fields exist)
  hex_gold <- hex |> dplyr::filter(!is.na(gold_label))

  popup_html <- ~ paste0(
    "<b>", GRID_ID, "</b><br/>",
    "Action: ", as.character(action_label), "<br/>",
    "WFE: ", WFE_CAT, " · People: ", PEOPLE_CAT, "<br/>",
    "Goldilocks: ",
    ifelse(is.na(gold_label), "Outside band", as.character(gold_label)),
    EVT_TOP3_HTML,
    "<br/>People-first urgency: ", urgency_0_100, " (0–100; 100 = highest in this map)"
  )

  bb <- st_bbox(hex)
  # Tiny margin only — avoid zooming out past the AOI
  pad_x <- (bb[["xmax"]] - bb[["xmin"]]) * 0.01
  pad_y <- (bb[["ymax"]] - bb[["ymin"]]) * 0.01
  west <- bb[["xmin"]] - pad_x
  south <- bb[["ymin"]] - pad_y
  east <- bb[["xmax"]] + pad_x
  north <- bb[["ymax"]] + pad_y

  leaflet(
    height = "85vh",
    width = "100%",
    options = leafletOptions(zoomSnap = 0.25, zoomDelta = 0.5)
  ) |>
    addProviderTiles("CartoDB.Positron") |>
    fitBounds(west, south, east, north) |>
    addPolygons(
      data = hex,
      # Map colors by name (not palette index) so labels never drift
      fillColor = ~ unname(action_colors[as.character(action_label)]),
      fillOpacity = ~ ifelse(action_label == "Wetlands — assess locally", 0.32, 0.72),
      color = "#333333",
      weight = 0.25,
      opacity = 0.5,
      group = "Action class",
      popup = popup_html
    ) |>
    addPolygons(
      data = hex_gold,
      fillColor = ~ unname(gold_colors[as.character(gold_label)]),
      fillOpacity = 0.85,
      color = "#145a32",
      weight = 0.6,
      opacity = 0.8,
      group = "Goldilocks",
      popup = popup_html
    ) |>
    addPolygons(
      data = hex[hex$is_wetland, ],
      fillColor = "#3d42ba",
      fillOpacity = 0.55,
      color = "#1a1a6e",
      weight = 0.4,
      opacity = 0.7,
      group = "Wetlands — assess locally",
      popup = popup_html
    ) |>
    addLegend(
      position = "bottomright",
      colors = unname(action_colors[action_levels]),
      labels = action_levels,
      title = "Action class",
      opacity = 0.9,
      group = "Action class"
    ) |>
    addLegend(
      position = "bottomright",
      colors = unname(gold_colors[gold_levels]),
      labels = gold_levels,
      title = "Goldilocks (people-first)",
      opacity = 0.9,
      group = "Goldilocks"
    ) |>
    addLegend(
      position = "bottomright",
      colors = "#3d42ba",
      labels = "Wetlands — assess locally",
      title = "Wetlands (toggle)",
      opacity = 0.9,
      group = "Wetlands — assess locally"
    ) |>
    addLayersControl(
      baseGroups = c("Action class", "Goldilocks"),
      overlayGroups = c("Wetlands — assess locally"),
      options = layersControlOptions(collapsed = FALSE)
    ) |>
    # Wetlands overlay off by default; turn on when viewing Goldilocks
    hideGroup("Wetlands — assess locally") |>
    htmlwidgets::onRender(sprintf(
      "
      function(el, x) {
        var map = this;
        var bounds = [[%f, %f], [%f, %f]];
        function legendTitle(leg) {
          var s = leg.querySelector('strong');
          return s ? s.textContent.trim() : '';
        }
        function syncLegends() {
          var legends = el.querySelectorAll('.legend');
          var inputs = el.querySelectorAll('.leaflet-control-layers-base input');
          var active = null;
          inputs.forEach(function(inp) {
            if (inp.checked) {
              active = inp.parentElement.textContent.trim();
            }
          });
          var overlayOn = false;
          el.querySelectorAll('.leaflet-control-layers-overlays input').forEach(function(inp) {
            if (inp.checked && inp.parentElement.textContent.indexOf('Wetlands') >= 0) {
              overlayOn = true;
            }
          });
          legends.forEach(function(leg) {
            var title = legendTitle(leg);
            var show = false;
            if (active === 'Action class' && title.indexOf('Action') === 0) show = true;
            if (active === 'Goldilocks' && title.indexOf('Goldilocks') === 0) show = true;
            if (overlayOn && title.indexOf('Wetlands') === 0) show = true;
            leg.style.display = show ? 'block' : 'none';
          });
        }
        function syncWetlandsOverlay() {
          var inputs = el.querySelectorAll('.leaflet-control-layers-base input');
          var active = null;
          inputs.forEach(function(inp) {
            if (inp.checked) active = inp.parentElement.textContent.trim();
          });
          // Prefer wetlands toggle with Goldilocks view
          var overlayInputs = el.querySelectorAll('.leaflet-control-layers-overlays input');
          overlayInputs.forEach(function(inp) {
            if (inp.parentElement.textContent.indexOf('Wetlands') < 0) return;
            var wantOn = (active === 'Goldilocks');
            if (inp.checked !== wantOn) inp.click();
          });
          syncLegends();
        }
        function fitAOI() {
          map.invalidateSize(false);
          map.fitBounds(bounds, { padding: [10, 10], animate: false });
        }
        map.on('baselayerchange', function() {
          syncWetlandsOverlay();
        });
        map.on('overlayadd overlayremove', syncLegends);
        setTimeout(function() {
          syncWetlandsOverlay();
          syncLegends();
        }, 50);
        var nFit = 0;
        function fitOnce() {
          fitAOI();
          nFit += 1;
        }
        fitOnce();
        setTimeout(fitOnce, 150);
        setTimeout(fitOnce, 500);
        setTimeout(fitOnce, 1200);
        if (window.ResizeObserver) {
          var ro = new ResizeObserver(function() {
            if (nFit < 8) fitOnce();
            else ro.disconnect();
          });
          ro.observe(el);
        }
      }
      ",
      south, west, north, east
    ))
}
```

**Action class** = what kind of conversation is proposed for that hex.

**Goldilocks** = which hexes to talk about first under a people-first lens (top 5% / 10% / 15% of actionable hexes across the AOI). Only in-band hexes are drawn on this view. **Wetlands — assess locally** are left out of those bands but can be toggled as a separate overlay.
