---
author: "Randy Swaty"
title: "Actions and Priority Areas: Fire Adapted Ecosystems Focus"
subtitle: "Setting the stage for conversation and deeper dives"
format:
  html:
    toc: false
    page-layout: full
    css: styles.css
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
library(sf)
library(leaflet)
library(dplyr)

hex_path <- "../outputs/hex/faa_hex_scores.geojson"
has_hex <- file.exists(hex_path)

action_levels <- c(
  "Value to protect from fire",
  "Treat fire risk for people",
  "Ecosystem health focus",
  "Wetlands — assess locally",
  "Defer and Monitor"
)
action_colors <- c(
  "Value to protect from fire" = "#ba620f",
  "Treat fire risk for people" = "#5c184b",
  "Ecosystem health focus" = "#0f451d",
  "Wetlands — assess locally" = "#b0b6f0",  # lighter than other actions
  "Defer and Monitor" = "#cfedca"
)

# Goldilocks heat: all hexes, white → dark purple by percentile rank of urgency.
gold_heat_colors <- c(
  "#ffffff", "#f2eef6", "#e0d4e7", "#c9b0d4",
  "#a67cba", "#7a3f96", "#542078", "#2d004b"
)
```


The Northwoods Region has no shortage of fire adapted ecosystems.  We have calculated that at least 4.6M acres are in moderate to high Wildfire Exposure Risk categories (see [initial exploration (Risk tab)](https://rswaty.github.io/northwoods/)). In this dashboard I work to prioritize what to do where with aggressive filtering, housing density data and draft "Action" assignments in hopes of getting at actionable areas. I am trying to find the middle ground, aka "The Goldilocks" range of areas-a stretch while also being doable.  **These maps are not meant to be the final answer** but instead guides of where to dig deeper.  Further, the 'Defer/Monitor' hexagons are not mean to be ignored, and may well have fire-focused risks to address.


You can toggle **Action class**, **Goldilocks**, **Fuel change (FDist)**, or **Basemap only** in the top-right panel. Goldilocks ranks **actionable hexes only** (not defer / wetlands) white → dark purple by urgency; defer and wetlands stay grey. Fuel change is its own view (brown = add, green = remove). Choose **Basemap only** to hide hexes and explore the map underneath. For more information also see [Context](context.qmd) and [Methods](methods.qmd).

```{r}
#| label: map
#| fig-height: 9
if (!has_hex) {
  cat(
    "**No hex output yet.** Run the Pro pipeline through `05_export_hex_geojson.py`, ",
    "commit `outputs/hex/faa_hex_scores.geojson`, then re-render this dashboard.\n"
  )
} else {
  hex <- st_read(hex_path, quiet = TRUE) |>
    mutate(
      action_label = case_match(
        ACTION_CLASS,
        "value_to_protect_from_fire" ~ "Value to protect from fire",
        "protect_from_fire" ~ "Value to protect from fire",  # legacy id
        "treat_fire_risk_for_people" ~ "Treat fire risk for people",
        "ecosystem_health_focus" ~ "Ecosystem health focus",
        "wetlands_assess_locally" ~ "Wetlands — assess locally",
        "defer_monitor" ~ "Defer and Monitor",
        .default = "Defer and Monitor"
      ),
      action_label = factor(action_label, levels = action_levels),
      is_wetland = ACTION_CLASS == "wetlands_assess_locally"
    )

  # Explicit levels= so leaflet/scales does not alphabetize and scramble colors.
  action_pal <- colorFactor(
    palette = unname(action_colors[action_levels]),
    domain = action_levels,
    levels = action_levels,
    ordered = TRUE,
    na.color = "#9D9D9D"
  )

  # Goldilocks rank among actionable hexes only (not defer / wetlands).
  # Homes-heavy defer towns (e.g. BR-33) must not outrank real action hexes.
  gold_rank_gamma <- 1.75
  hex <- hex |>
    mutate(
      SCORE_PEOPLE = as.numeric(SCORE_PEOPLE),
      gold_eligible = !ACTION_CLASS %in% c(
        "defer_monitor", "wetlands_assess_locally"
      ),
      FDIST_FUEL_DELTA = as.numeric(FDIST_FUEL_DELTA),
      fdist_dominate = dplyr::case_when(
        is.na(FDIST_FUEL_DELTA) ~ "n/a",
        FDIST_FUEL_DELTA >= 0.05 ~ "dominated by fuel add",
        FDIST_FUEL_DELTA <= -0.05 ~ "dominated by fuel remove",
        TRUE ~ "mixed / near neutral"
      )
    )

  # Percentile within actionable pool, then mild gamma; ineligible → NA (white)
  urg <- rep(NA_real_, nrow(hex))
  if (any(hex$gold_eligible)) {
    pr <- dplyr::percent_rank(hex$SCORE_PEOPLE[hex$gold_eligible])
    urg[hex$gold_eligible] <- round(100 * (pr^gold_rank_gamma), 0)
  }
  hex$urgency_0_100 <- urg

  gold_pal <- colorNumeric(
    palette = gold_heat_colors,
    domain = c(0, 100),
    na.color = "#f7f7f7"
  )

  # FDist: brown = fuel-add (+), green = fuel-remove (−)
  fdist_lim <- max(0.25, max(abs(hex$FDIST_FUEL_DELTA), na.rm = TRUE))
  fdist_pal <- colorNumeric(
    palette = c("#1b7837", "#f5f5f5", "#8c510a"),
    domain = c(-fdist_lim, fdist_lim),
    na.color = "#cccccc"
  )

  # Precompute EVT top-3 popup block (area ranks from script 03)
  if (all(c("EVT_1_NAME", "EVT_1_PCT") %in% names(hex))) {
    fmt_one <- function(nm, pct, rank) {
      ifelse(
        is.na(nm) | nm == "",
        "",
        paste0(
          rank, ". ", nm, " (",
          ifelse(is.na(pct), "?", paste0(round(pct, 0), "%")),
          ")"
        )
      )
    }
    e1 <- fmt_one(hex$EVT_1_NAME, hex$EVT_1_PCT, 1)
    e2 <- if ("EVT_2_NAME" %in% names(hex)) {
      fmt_one(hex$EVT_2_NAME, hex$EVT_2_PCT, 2)
    } else {
      rep("", nrow(hex))
    }
    e3 <- if ("EVT_3_NAME" %in% names(hex)) {
      fmt_one(hex$EVT_3_NAME, hex$EVT_3_PCT, 3)
    } else {
      rep("", nrow(hex))
    }
    hex$EVT_TOP3_HTML <- vapply(seq_len(nrow(hex)), function(i) {
      parts <- c(e1[i], e2[i], e3[i])
      parts <- parts[nzchar(parts)]
      if (!length(parts)) {
        return("")
      }
      paste0(
        "<br/><b>EVT (top 3 by area)</b><br/>",
        paste(parts, collapse = "<br/>"),
        "<br/><br/><i>Wildfire Exposure Risk is blended ember transmission across the hex fuel mosaic (Beverly-style), not majority EVT alone.</i>"
      )
    }, character(1))
  } else if ("EVT_NAME" %in% names(hex)) {
    hex$EVT_TOP3_HTML <- ifelse(
      is.na(hex$EVT_NAME) | hex$EVT_NAME == "",
      "",
      paste0("<br/>EVT: ", hex$EVT_NAME)
    )
  } else {
    hex$EVT_TOP3_HTML <- ""
  }

  # Keep wetland subset after popup fields exist
  hex_wetland <- hex[hex$is_wetland, ]

  # Ordinal suffix for popup ("2nd" not "2th")
  ordinal <- function(n) {
    n <- as.integer(n)
    ifelse(
      is.na(n),
      "n/a",
      paste0(
        n,
        ifelse(
          (n %% 100) %in% c(11L, 12L, 13L),
          "th",
          c("th", "st", "nd", "rd", "th", "th", "th", "th", "th", "th")[
            (n %% 10) + 1L
          ]
        )
      )
    )
  }

  popup_html <- ~ paste0(
    "<b>", GRID_ID, "</b><br/>",
    "Action: ", as.character(action_label), "<br/>",
    "Wildfire Exposure Risk: ", WFE_CAT, " · People: ", PEOPLE_CAT, "<br/>",
    "Goldilocks rank: ",
    ifelse(
      !gold_eligible,
      "n/a (defer / wetlands — not in Goldilocks pool)",
      paste0(ordinal(urgency_0_100), " percentile among actionable hexes")
    ),
    "<br/>Fuel change (FDist): ",
    ifelse(
      is.na(FDIST_FUEL_DELTA),
      "n/a",
      paste0(
        round(FDIST_FUEL_DELTA, 3),
        " — ",
        fdist_dominate
      )
    ),
    EVT_TOP3_HTML,
    "<br/>Color = rank among actionable hexes only (defer/wetlands grey)"
  )

  bb <- st_bbox(hex)
  # Tiny margin only — avoid zooming out past the AOI
  pad_x <- (bb[["xmax"]] - bb[["xmin"]]) * 0.01
  pad_y <- (bb[["ymax"]] - bb[["ymin"]]) * 0.01
  west <- bb[["xmin"]] - pad_x
  south <- bb[["ymin"]] - pad_y
  east <- bb[["xmax"]] + pad_x
  north <- bb[["ymax"]] + pad_y

  leaflet(
    height = "85vh",
    width = "100%",
    options = leafletOptions(zoomSnap = 0.25, zoomDelta = 0.5)
  ) |>
    addProviderTiles("CartoDB.Positron") |>
    fitBounds(west, south, east, north) |>
    addPolygons(
      data = hex,
      # Map colors by name (not palette index) so labels never drift
      fillColor = ~ unname(action_colors[as.character(action_label)]),
      fillOpacity = ~ ifelse(action_label == "Wetlands — assess locally", 0.32, 0.72),
      color = "#333333",
      weight = 0.1,
      opacity = 0.35,
      group = "Action class",
      popup = popup_html
    ) |>
    addPolygons(
      data = hex,
      fillColor = ~ gold_pal(urgency_0_100),
      fillOpacity = 0.78,
      color = "#666666",
      weight = 0.1,
      opacity = 0.3,
      group = "Goldilocks",
      popup = popup_html
    ) |>
    addPolygons(
      data = hex,
      fillColor = ~ fdist_pal(FDIST_FUEL_DELTA),
      fillOpacity = 0.78,
      color = "#333333",
      weight = 0.1,
      opacity = 0.35,
      group = "Fuel change (FDist)",
      popup = popup_html
    ) |>
    # Invisible layer so "Basemap only" can be a selectable base group
    addPolygons(
      data = hex[1, ],
      fillOpacity = 0,
      opacity = 0,
      weight = 0,
      group = "Basemap only",
      options = pathOptions(clickable = FALSE)
    ) |>
    addPolygons(
      data = hex_wetland,
      fillColor = "#3d42ba",
      fillOpacity = 0.55,
      color = "#1a1a6e",
      weight = 0.15,
      opacity = 0.5,
      group = "Wetlands — assess locally",
      popup = popup_html
    ) |>
    addLegend(
      position = "bottomright",
      colors = unname(action_colors[action_levels]),
      labels = action_levels,
      title = "Action class",
      opacity = 0.9,
      group = "Action class"
    ) |>
    addLegend(
      position = "bottomright",
      pal = gold_pal,
      values = c(0, 100),
      title = "Goldilocks rank (percentile)",
      opacity = 0.9,
      group = "Goldilocks"
    ) |>
    addLegend(
      position = "bottomright",
      pal = fdist_pal,
      values = c(-fdist_lim, fdist_lim),
      title = "Fuel change (− remove / + add)",
      opacity = 0.9,
      group = "Fuel change (FDist)"
    ) |>
    addLegend(
      position = "bottomright",
      colors = "#3d42ba",
      labels = "Wetlands — assess locally",
      title = "Wetlands (toggle)",
      opacity = 0.9,
      group = "Wetlands — assess locally"
    ) |>
    addLayersControl(
      baseGroups = c(
        "Action class",
        "Goldilocks",
        "Fuel change (FDist)",
        "Basemap only"
      ),
      overlayGroups = c("Wetlands — assess locally"),
      options = layersControlOptions(collapsed = FALSE)
    ) |>
    # Wetlands overlay off by default
    hideGroup("Wetlands — assess locally") |>
    htmlwidgets::onRender(sprintf(
      "
      function(el, x) {
        var map = this;
        var bounds = [[%f, %f], [%f, %f]];
        function legendTitle(leg) {
          var s = leg.querySelector('strong');
          return s ? s.textContent.trim() : '';
        }
        function syncLegends() {
          var legends = el.querySelectorAll('.legend');
          var inputs = el.querySelectorAll('.leaflet-control-layers-base input');
          var active = null;
          inputs.forEach(function(inp) {
            if (inp.checked) {
              active = inp.parentElement.textContent.trim();
            }
          });
          var overlayOn = false;
          el.querySelectorAll('.leaflet-control-layers-overlays input').forEach(function(inp) {
            if (inp.checked && inp.parentElement.textContent.indexOf('Wetlands') >= 0) {
              overlayOn = true;
            }
          });
          legends.forEach(function(leg) {
            var title = legendTitle(leg);
            var show = false;
            if (active === 'Action class' && title.indexOf('Action') === 0) show = true;
            if (active === 'Goldilocks' && title.indexOf('Goldilocks') === 0) show = true;
            if (active === 'Fuel change (FDist)' && title.indexOf('Fuel') === 0) show = true;
            if (overlayOn && title.indexOf('Wetlands') === 0) show = true;
            leg.style.display = show ? 'block' : 'none';
          });
        }
        function syncWetlandsOverlay() {
          var inputs = el.querySelectorAll('.leaflet-control-layers-base input');
          var active = null;
          inputs.forEach(function(inp) {
            if (inp.checked) active = inp.parentElement.textContent.trim();
          });
          var overlayInputs = el.querySelectorAll('.leaflet-control-layers-overlays input');
          overlayInputs.forEach(function(inp) {
            if (inp.parentElement.textContent.indexOf('Wetlands') < 0) return;
            // Keep wetlands off on basemap-only; otherwise leave user choice
            if (active === 'Basemap only' && inp.checked) inp.click();
          });
          syncLegends();
        }
        function fitAOI() {
          map.invalidateSize(false);
          map.fitBounds(bounds, { padding: [10, 10], animate: false });
        }
        map.on('baselayerchange', function() {
          syncWetlandsOverlay();
        });
        map.on('overlayadd overlayremove', syncLegends);
        setTimeout(function() {
          syncWetlandsOverlay();
          syncLegends();
        }, 50);
        var nFit = 0;
        function fitOnce() {
          fitAOI();
          nFit += 1;
        }
        fitOnce();
        setTimeout(fitOnce, 150);
        setTimeout(fitOnce, 500);
        setTimeout(fitOnce, 1200);
        if (window.ResizeObserver) {
          var ro = new ResizeObserver(function() {
            if (nFit < 8) fitOnce();
            else ro.disconnect();
          });
          ro.observe(el);
        }
      }
      ",
      south, west, north, east
    ))
}
```

**Action class** = what kind of conversation is proposed for that hex.

**Goldilocks** = white → dark purple **heat** on **actionable hexes only** (percentile rank of urgency). Defer and wetlands are grey — not ranked.

**Fuel change (FDist)** = recent fuel direction (brown = net add, green = net remove). Also **multiplies** the Goldilocks urgency score (add raises more than remove lowers).

**Basemap only** = hide hex fills so you can explore the basemap.
