---
title: "Next Gen FAA — Northwoods (draft)"
subtitle: "Fire Action Assessment — strategic screening only"
format:
  html:
    page-layout: full
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
library(sf)
library(leaflet)
library(dplyr)

hex_path <- "../outputs/hex/faa_hex_scores.geojson"
has_hex <- file.exists(hex_path)
```

Strategic screening map for northern MI, WI, and MN Arrowhead. Scores and action classes are produced in **ArcGIS Pro** (Python in `src/`); this page only displays pushed hex outputs. Default emphasis: **people-first** (`SCORE_PEOPLE` / Goldilocks). **PAD** is shown as context only (not used in scoring).

```{r}
#| label: map
#| fig-height: 7
if (!has_hex) {
  cat(
    "**No hex output yet.** Run the Pro pipeline through `05_export_hex_geojson.py`, ",
    "commit `outputs/hex/faa_hex_scores.geojson`, then re-render this dashboard.\n"
  )
} else {
  hex <- st_read(hex_path, quiet = TRUE)

  action_pal <- colorFactor(
    palette = c("#B279A2", "#E45756", "#54A24B", "#2CA6A4", "#9D9D9D"),
    domain = c(
      "protect_from_fire",
      "treat_fire_risk_for_people",
      "ecosystem_health_focus",
      "wetlands_assess_locally",
      "defer_monitor"
    ),
    na.color = "#9D9D9D"
  )

  pad_pal <- colorNumeric(
    palette = "Blues",
    domain = if ("PADUS_FRAC" %in% names(hex)) hex$PADUS_FRAC else c(0, 1),
    na.color = "#cccccc"
  )

  m <- leaflet(hex) |>
    addProviderTiles("CartoDB.Positron") |>
    addPolygons(
      fillColor = ~ if ("ACTION_CLASS" %in% names(hex)) {
        action_pal(ACTION_CLASS)
      } else {
        "#9D9D9D"
      },
      fillOpacity = 0.55,
      color = "#444444",
      weight = 0.3,
      group = "Action class",
      popup = ~ paste0(
        "<b>", if ("GRID_ID" %in% names(hex)) GRID_ID else "hex", "</b><br/>",
        if ("ACTION_CLASS" %in% names(hex)) paste0("Action: ", ACTION_CLASS, "<br/>") else "",
        if ("TREATMENT_HINT" %in% names(hex)) paste0("How: ", TREATMENT_HINT, "<br/>") else "",
        if ("GOLDILOCKS_PRIORITY" %in% names(hex)) paste0("Priority (0-3): ", GOLDILOCKS_PRIORITY, "<br/>") else "",
        if ("SCORE_PEOPLE" %in% names(hex)) paste0("People-first score: ", round(SCORE_PEOPLE, 3), "<br/>") else "",
        if ("PADUS_FRAC" %in% names(hex)) paste0("PAD frac (context): ", round(PADUS_FRAC, 3), "<br/>") else "",
        if ("FDIST_FUEL_DELTA" %in% names(hex)) paste0("FDist fuel delta: ", round(FDIST_FUEL_DELTA, 3), "<br/>") else "",
        if ("PINE_HEX" %in% names(hex)) paste0("Pine/barrens hex: ", PINE_HEX, "<br/>") else "",
        if ("EVT_NAME" %in% names(hex)) paste0("EVT: ", EVT_NAME, "<br/>") else "",
        if ("EVT_MAJORITY" %in% names(hex)) paste0("EVT code: ", EVT_MAJORITY, "<br/>") else "",
        if ("EVT_FIRE" %in% names(hex)) paste0("EVT FIRE (−1/0/1): ", EVT_FIRE) else ""
      )
    )

  if ("PADUS_FRAC" %in% names(hex)) {
    m <- m |>
      addPolygons(
        fillColor = ~ pad_pal(PADUS_FRAC),
        fillOpacity = 0.55,
        color = "#444444",
        weight = 0.2,
        group = "PAD context (GAP 1–3 frac)",
        popup = ~ paste0(
          "<b>", if ("GRID_ID" %in% names(hex)) GRID_ID else "hex", "</b><br/>",
          "PADUS_FRAC: ", round(PADUS_FRAC, 3), " (context only)"
        )
      ) |>
      addLegend(
        position = "bottomleft",
        pal = pad_pal,
        values = ~ PADUS_FRAC,
        title = "PAD frac",
        group = "PAD context (GAP 1–3 frac)"
      )
  }

  groups <- "Action class"
  if ("PADUS_FRAC" %in% names(hex)) {
    groups <- c(groups, "PAD context (GAP 1–3 frac)")
  }

  m <- m |>
    addLegend(
      position = "bottomright",
      pal = action_pal,
      values = ~ if ("ACTION_CLASS" %in% names(hex)) ACTION_CLASS else "defer_monitor",
      title = "Action class"
    ) |>
    addLayersControl(
      baseGroups = groups,
      options = layersControlOptions(collapsed = FALSE)
    )

  m
}
```

## Weight presets (from config)

Dashboard default emphasis is **people-first**. PAD is **context only** (layer toggle above; not in Goldilocks score). Fuel-add and pine/barrens flags feed actions after you re-run 03→05. Goldilocks ranks **actionable** hexes only (`defer_monitor` excluded); `GOLDILOCKS_PRIORITY` runs 0–3.
